Analyzing Geographic and Demographic Patterns of Homelessness

### 2024-01-25 outputs/paper/paper.qmd
updated R project name

This study is designed for an in-depth analysis of Toronto's shelter occupancy and capacity, utilizing data [@ShelterData] from the Shelter Management Information System (SMIS) database. While there are several papers on homelessness, this paper more focus on data-driven analyses that integrate both demographic and geographic dimensions within the context of urban shelter systems. Other datasets have often been limited either in data scope or timeframe, failing to provide a comprehensive picture that reflects the current realities of urban homelessness. This paper addresses this gap better by offering an updated view of the shelter system in Toronto.

@@

This research is based on a dataset sourced from the Shelter Support and Housing Administration (SSHA) division's Shelter Management Information System (SMIS) database, which is accessible through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto]. The specific dataset employed is named 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData]. This database acts as a detailed collection of daily entries related to active overnight shelter and associated services. It is the latest and most comprehensive dataset available for analyzing homelessness support services.

@@

1. While the 'Toronto Shelter System Flow' [@ShelterSystemFlow] data offers valuable insights into the flow of demography through the shelter system, it lacks comprehensive geographic information. This limitation significantly hinders the potential for a detailed analysis of geographic trends and patterns within the shelter system, an aspect critical to my research objectives.

2. The 'Daily Shelter Occupancy' [@DailyShelterOccupancy] dataset, although previously a proper source of information, has become less relevant due to its discontinuation. The most recent update for this dataset was recorded on March 10, 2022, therefore it is outdated for current analysis purposes. In contrast, the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset provides the most recent and comprehensive data available, making it a more suitable choice for analyzing up-to-date trends and developments in shelter usage and capacity.

@@

The latest analysis of shelter occupancy data reveals a marked upward trend in average occupancy rates, rounded up at the first decimal place, over 21 days from January 1st to January 21st in 2024 (see @fig-average-occupancy-rate-by-date). While daily fluctuations were observed, the general direction was a steady increase in the need for shelter services. However, to provide a more comprehensive understanding, several critical aspects need to be considered in the discussion:

@@

2. **Expanding Data Parameters**: Future studies should aim to include additional variables such as the program model, service type, and program area of shelters. This expansion will allow for a more comprehensive analysis, potentially uncovering patterns and trends that are not evident from occupancy rates alone. For example, understanding how different program models affect occupancy rates can inform more targeted and effective policy decisions and resource allocation.

In conclusion, while this study provides important insights into shelter occupancy trends, its scope and the recency of data are somewhat limited. Future research efforts should aim to overcome these limitations by incorporating more comprehensive and up-to-date data, thereby enriching the understanding of homelessness and informing more effective responses to this pressing social issue.

>>>>

This study is designed for an in-depth analysis of Toronto's shelter occupancy and capacity, utilizing data [@ShelterData] from the Shelter Management Information System (SMIS) database. While there are several papers on homelessness, this paper more focus on data-driven analyses that integrate both demographic and geographic dimensions within the context of urban shelter systems. Other datasets have often been limited either in data scope or timeframe, failing to provide a picture that reflects the current realities of urban homelessness. This paper addresses this gap better by offering an updated view of the shelter system in Toronto.

@@

This research is based on a dataset sourced from the Shelter Support and Housing Administration (SSHA) division's Shelter Management Information System (SMIS) database, which is accessible through the City of Toronto’s OpenDataToronto Library [@rOpenDataToronto]. The specific dataset employed is named 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData]. This database acts as a detailed collection of daily entries related to active overnight shelters and associated services.

@@

1. While the 'Toronto Shelter System Flow' [@ShelterSystemFlow] data offers valuable insights into the flow of demography through the shelter system, it lacks geographic information. This limitation significantly hinders the potential for a detailed analysis of geographic trends and patterns within the shelter system, an aspect critical to my research objectives.

2. The 'Daily Shelter Occupancy' [@DailyShelterOccupancy] dataset, although previously a proper source of information, has become less relevant due to its discontinuation. The most recent update for this dataset was recorded on March 10, 2022, therefore it is outdated for current analysis purposes. In contrast, the 'Daily Shelter & Overnight Service Occupancy & Capacity' [@ShelterData] dataset provides the most recent and thorough data available, making it a more suitable choice for analyzing up-to-date trends and developments in shelter usage and capacity.

@@

The latest analysis of shelter occupancy data reveals a marked upward trend in average occupancy rates, rounded up at the first decimal place, over 21 days from January 1st to January 21st in 2024 (see @fig-average-occupancy-rate-by-date). While daily fluctuations were observed, the general direction was a steady increase in the need for shelter services. However, to provide a more thorough understanding, several critical aspects need to be considered in the discussion:

@@

2. **Expanding Data Parameters**: Future studies should aim to include additional variables such as the program model, service type, and program area of shelters. This expansion will allow for a more thorough analysis, potentially uncovering patterns and trends that are not evident from occupancy rates alone. For example, understanding how different program models affect occupancy rates can inform more targeted and effective policy decisions and resource allocation.

In conclusion, while this study provides important insights into shelter occupancy trends, its scope and the recency of data are somewhat limited. Future research efforts should aim to overcome these limitations by incorporating more thorough and up-to-date data, thereby enriching the understanding of homelessness and informing more effective responses to this pressing social issue.
